---
title: "Simple spatial interaction model in R"
author: "Robin Lovelace"
date: "23/07/2014"
output: html_document
---

This document shows the construction of a very basic spatial interaction model in R. It builds on a previous [unconstrained model](https://github.com/Robinlovelace/twitter-leeds) with a new set of data. Before loading the data, we set-up R environment:

```{r, message=FALSE}
# Load packages and setup global parameters
x <- c("gdata", "rgeos", "sp") # packages we'll be using
lapply(x, require, character.only = T)
set.seed(0) # make probabilistic results reproducible
beta <- 0.3
```

## Loading the input data

The data are stored in in the stores.csv file - the `CENTRE_SIZE` variable has been
modified to avoid the disclosure of commercially sensitive information:

```{r}
d <- read.csv("data/stores.csv") # the 80 locations used
coords <- matrix(c(d$EAST, d$NORTH), ncol = 2) # coordinates of dests
d <- SpatialPointsDataFrame(coords, data = d) # convert to spatial data
d$pop <- rnorm(n = nrow(d), mean = 1000, sd = 200) # random population variable
```

## Generating the distance matrix
Because R sees the `d` object as a spatial point, we can produce a distance matrix between different points with a single line of code. The second line below creates
a new matrix of the same dimension, set to 0, for recording the flow:

```{r}
r <- gDistance(d, d, byid = T) / 1000
f <- r; f[] <- 0
```

## Run and plot the basic model

The only remaining task is to run the model, in a nested for loop over origins (i) and destinations (j). Note the omission of flows between origins and destinations that are the same with `n[-i]`:

```{r}
plot(d)
n <- 1:nrow(d)
for(i in n){
  for(j in n[-i]){
    f[i,j] <- d$pop[i] * d$CENTRE_SIZE[j] * exp(-beta * r[i,j])
    fw <- f[i,j]^0.3 / 10
    lines(c(d$EAST[i], d$EAST[j]), c(d$NORTH[i], d$NORTH[j]), lwd = fw)
  }  
}
```

## Further work

There are many ways we can refine and extend the above model.
In future posts we'll see how to add more parameters and allow
the population of each origin to determine the outward flow.
We'll also take a look at how to optimise the using the `optim()`
function.

Eventually, the aim is to build a fully operations [radiation model]()
and compare this with the classic gravity model presented in this post.
The oportunities for further work in this area are vast, so
watch this space for further work.


